### 2019-05-10 paper/main.tex
first draft done and added baseline_50

\usepackage{draftwatermark}
\SetWatermarkText{Draft}
\SetWatermarkColor[gray]{0.9}

@@

@@

% Links
\usepackage[colorlinks=true, citecolor=base08, linkcolor=base08, urlcolor=base08]{hyperref}

@@

\chapter*{Abstract}
    Lorem Ipsum

@@

\chapter{A hazy shade of winter}
%Simon and Garfunkel - A hazy shade of winter

@@

To combat mode collapse, \textcite{salimans_improved_2016} propose a technique called feature matching, where we change the objective of the generator to generate real data that matches the statistics of the real data. Since in the setting of a GAN showing the real data would be considered "cheating", we use the features of an intermediate layer of the discriminator. Specifically, we train the generator to match the expected value of those features \parencite{salimans_improved_2016}. Denoting $f(x)$ as the activations on an intermediate layer in the discriminator, the new objective for the generator becomes
\begin{equation}
    \|\mathbb{E}_{x~p_{data}} f(x) - \mathbb{E}_{z~p_z (z)} f(G(z))\|_2^2
\end{equation}
which is commonly known as the squared error. However, to make the re-implementation of C-RNN-GAN true to the original, Mean Square Error (MSE) was used as the loss function when using feature matching.

@@

To establish the degree to which the music of a Reptile-based model is comparable to (1) a model trained on whole dataset and (2) songs from the dataset, a total of three models have been trained (see Table \ref{tab:params} for a list of training parameters). Training of the baseline finished after 27 hours, 9 hours for the Reptile adaptation of Performance-RNN, and 19 hours for the Reptile adaptation of C-RNN-GAN.

@@

Having said that, the landscape changes a bit when looking at NDB, our other evaluation metric. With seven to ten bins statistically different (out of 20), our models seem to display a great deal of disparity form the test samples, even though they somewhat agree between themselves - in other words, they don't seem to be very different from each other.

With these result we have gathered enough information to answer our research questions, where we could claim that our C-RNN-GAN implementation produces some of the most best results (as it scores best on 3/4 domain-specific metrics). However, a deep-dive into the performance at training time paints a whole different picture.

\begin{landscape}
\begin{table}[t]
    \centering
\begin{tabular}{l|l|l|l|l|l|l}
    Model             & Temperature & Polyphony       & Repetitions       & Tone Span        & Scale Consistency & NDB     \\ \hline
\textit{Test Set} & N/A         & \textit{0.4915} & \textit{814.3280} & \textit{66.9440} & \textit{0.6827}   & N/A     \\
Baseline          & 0.8         & 0.5707          & \textbf{\color{base0D}388.0160}    & 78.4480          & 0.6770            & 11      \\
Baseline          & 0.9         & 0.6054          & 373.0560          & 81.3120          & 0.6747            & 10      \\
Baseline          & 1.0         & 0.6112          & 347.3600          & 83.0560          & 0.6728            & 10      \\
Baseline          & 1.1         & 0.6292          & 334.4160          & 84.4400          & 0.6722            & \textbf{\color{base0D}7} \\
Baseline          & 1.2         & 0.6060          & 316.8160          & 85.0480          & 0.6726            & 9       \\
Performance-RNN   & 0.8         & 0.5684          & 377.8240          & 78.8400          & 0.6760            & 9       \\
Performance-RNN   & 0.9         & 0.5852          & 364.9680          & 81.4880          & 0.6757            & 8       \\
Performance-RNN   & 1.0         & 0.6219          & 354.0800          & 83.2400          & 0.6730            & 9       \\
Performance-RNN   & 1.1         & 0.6342          & 343.2240          & 84.4640          & 0.6721            & 9       \\
Performance-RNN   & 1.2         & 0.6061          & 312.2480          & 85.2400          & 0.6723            & 10      \\
C-RNN-GAN         & 0.8         & \textbf{\color{base0D}0.5505}    & 369.5760          & \textbf{\color{base0D}77.7920}    & \textbf{\color{base0D}0.6776}      & 9       \\
C-RNN-GAN         & 0.9         & 0.5940          & 370.5040          & 81.4320          & 0.6749            & 9       \\
C-RNN-GAN         & 1.0         & 0.6060          & 348.3520          & 83.2640          & 0.6733            & 8       \\
C-RNN-GAN         & 1.1         & 0.6343          & 340.4160          & 84.6800          & 0.6730            & 11      \\
C-RNN-GAN         & 1.2         & 0.6057          & 313.6080          & 85.0080          & 0.6726            & 10     
\end{tabular}
\caption{Results of the evaluation process of the generated sample. For comparison, relevant statistics of the test samples have been included. The values which are closest to the test sample are in bold and highlighted in green.}
\label{tab:results}
\end{table}
\end{landscape}

@@

    \begin{subfigure}{0.5\linewidth}
        \centering
        \input{images/baseline.tex}
        \caption{Baseline loss at training time}
        \label{fig:loss_baseline}
    \end{subfigure}
    \begin{subfigure}{0.5\linewidth}
        \centering
        \input{images/performance_rnn_1_5.tex}
        \caption{Performance-RNN loss at training time}
        \label{fig:loss_performancernn}
    \end{subfigure}

@@

    \begin{subfigure}{\linewidth}
        \centering
        \input{images/c_rnn_gan_1_5.tex}
        \caption{C-RNN-GAN loss at training time}
        \label{fig:loss_crnngan}
    \end{subfigure}
    \caption{Training losses of the three models featured in this paper. Note that the C-RNN-GAN reimplementation has a different loss function than the other two. Also to note, the loss of C-RNN-GAN is zero for both the generator and discriminator throughout most of the training time.}
    \label{fig:loss}
\end{figure}

When looking at Figure \ref{fig:loss} we gain an insight on how well the training has gone. First and foremost, the loss of Performance-RNN is as one would expect: the model starts somewhat poorly, and then incrementally tunes itself to perform better (Figure \ref{fig:loss_performancernn}). The diminishing up-and-down motions are most likely indicative of the optimizer's carving its path to convergence to a (hopefully) global minima.

@@

Finally, we have the losses of the GAN, a bewildering tableau with more questions than answers (Figure \ref{fig:loss_crnngan}). Firstly, it should be mentioned that GANs in and of themselves are somewhat finicky to train. Secondly, in a normal game of minimax, the loss of the generator and that of the discriminator should alternate adversarially: when one goes up, the other goes down. Here, they move together, to the point where both models reach a loss of zero - meaning that the generator is always able to fake the discriminator and the discriminator is always able to distinguish the fake samples from the real samples. This is likely due to the implementation of feature matching where instead of being trained in a competition, they are trained independently. However, this is the only explanation that I can come up with, and without other data with respect to the way the training progresses, it is difficult to say with certainty.

Having said that, whichever is the reason, does not change the evidence to the study's research questions. Respectively, our data suggests that models trained under Reptile are directly competitive to the baseline (a model trained on the entire dataset). However, this does not mean that they are "good enough". When compared to our test samples, the data points generated by the Reptile models fail to reach the same level of quality, a limitation likely imposed by the number of training iterations.

@@

@@

\printbibliography

>>>>

%\usepackage{draftwatermark}
%\SetWatermarkText{Draft}
%\SetWatermarkColor[gray]{0.9}

@@

% Links
\usepackage{url}
\def\UrlBreaks{\do\/\do-}
\setcounter{biburllcpenalty}{7000}
\setcounter{biburlucpenalty}{8000}

\usepackage[colorlinks=true, citecolor=base08, linkcolor=base08, urlcolor=base08]{hyperref}

@@

@@

%TC:ignore
\chapter*{Abstract}
Experiments in Deep Learning are marked by scant explorations outside of the classification task, and even more so when trying to overcome some of the limitations imposed by gradient-based optimization algorithms (such as the uncertainty of convergence when dealing with non-convexity). With the resurgence of generative models and recent advancements in meta-learning we are faced two exciting ares of research, the intersection of which could be a promising area when investing general-purpose A.I. systems. Inspired by \textcite{clouatre_figr_2019} and \textcite{zhang_metagan_2018}, this paper proposed a new line of inquiry where these two fields could be useful specifically, in the musical domain. As such, the paper sought to establish the extent to which meta-learned generative models are comparable to a model trained under normal circumstances, and to real musical performances.

Three models have been trained to generate music akin to the performances of expert musicians participating at the International Piano-e-Competition. Two of these models are models come from the literature and followed a 1-shot 5-way training with Reptile, a recent meta-learning algorithm. Afterwards they were evaluated against the baseline model (a LSTM network consisting of one layer with 256 units), and the test samples from the MAESTRO dataset.

When compared to the baseline, the two Reptile-based models have show to be quite competitive however, they failed to attain the same level of mastery when compared to the test samples. However, it is hypothesized that more time involved into training them could narrow the gap between the models and the real musical performances. Thus, the results of these experiments solidify the intuition that meta-learning algorithms are a viable alternative to traditional training methods, in the context of generative models.

%TC:endignore

@@

\chapter{A hazy shade of winter} \label{chap:intro}
%Simon and Garfunkel - A hazy shade of winter

@@

To combat mode collapse, \textcite{salimans_improved_2016} propose a technique called feature matching, where we change the objective of the generator to generate real data that matches the statistics of the real data. Specifically, we train the generator to match the expected value of those features \parencite{salimans_improved_2016}. However, this measure was not implemented in this paper as early tests caused both the loss of both the generator and discriminator to drop at zero, causing both models to stop learning.

@@

To establish the degree to which the music of a Reptile-based model is comparable to (1) a model trained on whole dataset and (2) songs from the dataset, a total of three models have been trained (see Table \ref{tab:params} for a list of training parameters). Training of the baseline finished after 27 hours, 9 hours for the Reptile adaptation of Performance-RNN, and 13 hours for the Reptile adaptation of C-RNN-GAN.

@@

Having said that, the landscape changes a bit when looking at NDB, our other evaluation metric. With seven to ten bins statistically different (out of 20), our models seem to display a great deal of disparity form the test samples, even though they somewhat agree between themselves - i.e., they don't seem to be very different from each other.

With these result we have gathered enough information to answer our research questions, and claim that the baseline still produces some of the best results (as it scores best on 4/5 metrics). However, a deep-dive into the performance at training time would shed some more light to these results.

\begin{landscape}
\begin{table}[t]
    \centering
\begin{tabular}{l|l|l|l|l|l|l}
    Model             & Temperature & Polyphony (\%)  & Repetitions       & Tone Span        & Scale Consistency (\%) & NDB     \\ \hline
\textit{Test Set} & N/A         & \textit{49.15} & \textit{814.33} & \textit{66.9440} & \textit{68.27}   & N/A     \\
Baseline          & 0.8         & 57.07          & \textbf{\color{base0D}388.02}    & \textbf{\color{base0D}78.45}          & \textbf{\color{base0D}67.70}            & 11      \\
Baseline          & 0.9         & 60.54          & 373.06          & 81.31          & 67.47            & 10      \\
Baseline          & 1.0         & 61.12          & 347.36          & 83.06          & 67.28            & 10      \\
Baseline          & 1.1         & 62.92          & 334.42          & 84.44          & 67.22            & \textbf{\color{base0D}7} \\
Baseline          & 1.2         & 60.60          & 316.82          & 85.05          & 67.26            & 9       \\
Performance-RNN   & 0.8         & 56.84          & \textbf{\color{base0A}377.82}          & 78.84          & 67.60            & 9       \\
Performance-RNN   & 0.9         & 58.52          & 364.97          & 81.49          & 67.57            & 8       \\
Performance-RNN   & 1.0         & 62.19          & 354.08          & 83.24          & 67.30            & 9       \\
Performance-RNN   & 1.1         & 63.42          & 343.22          & 84.46          & 67.21            & 9       \\
Performance-RNN   & 1.2         & 60.61          & 312.25          & 85.24          & 67.23            & 10      \\
C-RNN-GAN         & 0.8         & \textbf{\color{base0E}56.59}    & 375.25          & \textbf{\color{base0A}78.82}    & \textbf{\color{base0A}67.63}      & 9       \\
C-RNN-GAN         & 0.9         & 58.41          & 365.19          & 81.22          & 67.50            & 9       \\
C-RNN-GAN         & 1.0         & 61.25          & 351.02          & 83.00          & 67.40            & \textbf{\color{base0A}8}       \\
C-RNN-GAN         & 1.1         & 63.49          & 338.39          & 84.66          & 67.29            & 11      \\
C-RNN-GAN         & 1.2         & 60.57          & 316.35          & 85.30          & 67.19            & 10     
\end{tabular}
\caption{Results of the evaluation process of the generated sample. For comparison, relevant statistics of the test samples have been included. Important values are in bold. The values which are closest to the test sample are in green. Values which are closest to the baseline's best score are in red. Values which are both close to the baseline's best and the test sample are in purple.}
\label{tab:results}
\end{table}
\end{landscape}

@@

    \begin{subfigure}{0.5\linewidth}
        \centering
        \input{images/performance_rnn_1_5.tex}
        \caption{Performance-RNN loss at training time}
        \label{fig:loss_performancernn}
    \end{subfigure}
    \begin{subfigure}{0.5\linewidth}
        \centering
        \input{images/c_rnn_gan_1_5.tex}
        \caption{C-RNN-GAN loss at training time}
        \label{fig:loss_crnngan}
    \end{subfigure}

@@

    \begin{subfigure}{\linewidth}
        \centering
        \input{images/baseline.tex}
        \caption{Baseline loss at training time}
        \label{fig:loss_baseline}
    \end{subfigure}
    \caption{Training losses of the three models featured in this paper. Note that the C-RNN-GAN reimplementation has a different loss function than the other two. Also to note, the loss of C-RNN-GAN is zero for both the generator and discriminator throughout most of the training time.}
    \label{fig:loss}
\end{figure}

When looking at Figure \ref{fig:loss} we gain an insight on how well the training has gone. First and foremost, the loss of Performance-RNN is as one would expect: the model starts somewhat poorly, and then incrementally tunes itself to perform better (Figure \ref{fig:loss_performancernn}). Then, we have the losses of the GAN (Figure \ref{fig:loss_crnngan}) which portrays the adversarial training the generator and discriminator. Although on a first glance one might think that the loss is subject to high variance, it should be noted that a different function was used in this scenario and that the plot is on a much smaller scale than all others. Nevertheless, it does hint that the generator was slowly getting better at tricking the discriminator.

@@

    \newlength\heatheight
    \newlength\heatwidth
    \setlength\heatheight{9.5cm}
    \setlength\heatwidth{11.5cm}
\begin{figure}
        \begin{tabular}{c}
            \input{images/baseline/baseline-gen-.tex} \\
            \input{images/baseline/baseline-acts-.tex} \\
    \end{tabular}
    \caption{Baseline generated sample (top) and neuron activations for the respective event (bottom). The generated sample has a size of 200. The intensity of the color indicates the strength of the activation.}
    \label{fig:baseline_activations}
\end{figure}

\begin{figure}
        \begin{tabular}{c}
            \input{images/performance_rnn/performance_rnn_1_5-gen-.tex} \\
            \input{images/performance_rnn/performance_rnn_1_5-acts-.tex} \\
    \end{tabular}
    \caption{Performance-RNN generated sample (top) and neuron activations for the respective event (bottom). The generated sample has a size of 200. The intensity of the color indicates the strength of the activation.}
    \label{fig:performancernn_activations}
\end{figure}

\begin{figure}
        \begin{tabular}{c}
            \input{images/c_rnn_gan/c_rnn_gan_1_5-gen-.tex} \\
            \input{images/c_rnn_gan/c_rnn_gan_1_5-acts-.tex} \\
    \end{tabular}
    \caption{C-RNN-GAN generated sample (top) and neuron activations for the respective event (bottom). The generated sample has a size of 200. The intensity of the color indicates the strength of the activation.}
    \label{fig:crnngan_activations}
\end{figure}

\begin{figure}[t]
    \setlength\heatwidth{6.25cm}
    \begin{subfigure}[b]{0.5\linewidth}
        \setlength\heatheight{8cm}
        \input{images/baseline/baseline-acts-transposed.tex}
    \end{subfigure}
    \begin{subfigure}[b]{0.5\linewidth}
        \setlength\heatheight{14.5cm}
        \input{images/performance_rnn/performance_rnn_1_5-acts-transposed.tex}
    \end{subfigure}
    \caption{At scale neuron activations of the baseline (left) and Performance-RNN (right)}
    \label{fig:acts_scaled}
\end{figure}

To add another layer of validation to the results in Table \ref{tab:results}, Softmax activations of the last (or only) LSTM layer for each of our models was plotted on a heatmap. These activations take place when the model is tasked with generating a song. When accounting for the fact that the scale of the X-axis of the neuron activation plot differs for each model (because each one has a different hidden size), a couple of things become concrete. Firstly, that the generating process of Performance-RNN (Figure \ref{fig:performancernn_activations}) is sparse but still similar to that of the baseline (Figure \ref{fig:baseline_activations}). To better illustrate this, Figure \ref{fig:acts_scaled} shows the two applications on the same scale. Secondly, that C-RNN-GAN (Figure \ref{fig:crnngan_activations}) still has a long way to go until it can be considered adequate at generating a song; in other words, C-RNN-GAN generates songs mostly at random, claim enforced by the fact that almost all neurons are activated and that the pattern of generating a sequence has no order.

With these in mind, the data suggests that models trained under Reptile can be competitive with a model that has been trained under regular circumstances. However, this does not mean that they are all "good enough". When compared to our test samples, the songs generated by the Reptile models fail to reach the same level of quality, a limitation likely imposed by the number of training iterations.

@@

From the lens of mainstream media, we are at a point where we will soon create what is commonly known as "Artificial Intelligence" \parencite{levy_ai_2010}. This claim, of course, ignores the popularity cycles of conectionism dating back to the fifties \parencite[e.g.,][]{rosenblatt_perceptron_1957}, or the over-inflated expectations for autonomous human-made systems that were made ever since \parencite{marcus_is_2012}. However, even after suffering waves of funding-cuts, masses of criticism, and regressions into obscurity, develpments were still being made \parencite{knight_ai_2016, minsky_perceptrons_1988} - albeit at a smaller pace. With the reemergence of ANNs, we are now seeing even stronger developments towards creating general-purpose Artificial Intelligence \parencite[e.g.,][]{vinyals_starcraft_2017}, but that should not cause any relaxation or distress; "we are not there yet".

In Chapter \ref{chap:intro}, I pointed out a couple of shortcomings that current ANN systems have (reliance on massive amounts of data, and inability to perform separate tasks) along with two exciting trends in Deep Learning: the resurgence of generative networks and successfully infusing a system with meta-learning capabilities). Acknowledging the initial attempts at intersecting these two fields, this paper proposed expanding those efforts to the musical domain. It set to develop two generative models by modelling the training procedure to be more akin to that of a human learning, and evaluated their performances against a baseline proposed by \textcite{larochelle_few-shot_2017} and their resemblance to real samples. Concretely, Performance-RNN \parencite{oore_this_2018} and C-RNN-GAN \parencite{mogren_c-rnn-gan_2016} were trained under the Reptile learning framework \parencite{nichol_first-order_2018}, and then evaluated against a LSTM network with one layer and 256 hidden units as well as expert piano performances extracted from the MAESTRO dataset \parencite{hawthorne_enabling_2018}.

In general, the baseline recorded the best performance. However, this does not change the fact that it was closely followed by the Reptile models. More appropriately is to point that, in spite of being trained to generate music using a handful of examples, the Reptile models closely matched the performance of the baseline. This especially holds true for this paper's reimplementation of Performance-RNN whose neuron activations show that it has indeed learned how to generate music, even though its samples scored somewhere in the middle. The adversarial model showed that it is still somewhere in the early stages of learning, which would explain why almost all of its neurons were fired but, the fact that it performed reasonably in its evaluation should not be ignored either. Having said these, I say with confidence that when tasked with making music, Reptile-based generative models can be quite competitive.

As well as they perform, both few-shot models failed to produce samples that could be passed as real. However, neither did the baseline, which would indicate that with more time invested into training, all three models could generate samples that are both novel and high fidelity.

Which leads to one of the first limitations of this paper, namely the training procedure. Optimally, all models should be trained for over 1,000 epochs however, time and hardware constrains meant taking some shortcuts. This constraint is especially seen in the case of the baseline which has been trained with a batch size 32 times larger than the Reptile models (to drive down the training time from 250 hours to 27). In addition, it is likely that the training of the GAN model suffered from the fact that a measure for combating mode-collapse was not implemented. Had it been done so, then it's possible that the model would have had better activations and consequently, better samples.

Next, the generating process from which we sample with temperature to generate the next likely event is a somewhat greedy process \parencite{holtzman_curious_2019}. Assuming that the probability distribution of how humans make music borrows some of its characteristics from how we create language - which is likely since both tasks share a significant portion of brain activations \parencite{brown_music_2006} -, then greedy sampling or beam search (another sampling method) would completely fail to generate human-like music. Specifically, they are not ideal and nor adequate for a high entropy task \parencite{holtzman_curious_2019}. This could explain why the models in this paper lacked the concept of motif. Curiously enough, two recent solutions for sampling tokens have been shown to be better candidates: top-$k$ sampling \parencite{fan_hierarchical_2018} and nucleus sampling \parencite[also known as top-$p$ sampling;][]{holtzman_curious_2019}. Future work could incorporate either of these proposed sampling methods and evaluate them against the more traditional sampling methods in the context of generating music with a meta-learning algorithm.

Which brings me to a final limitation. Performance-RNN and C-RNN-GAN are using Reptile, a first-order meta-learning algorithm. \textcite{nichol_first-order_2018} have provided clear evidence that this algorithm is more stable than MAML's first-order variant. However, their results also point that vanilla MAML is still superior. Having said that, these algorithms encompass only one approach to meta-learning (initialisation-based learning) whereas hallucination or distance metric methods could paint a different picture \parencite{chen_closer_2018}. Thus, future work could experiment not only with MAML, but also other types of algorithms.

Nevertheless, the results in this paper are intriguing in and of themselves. They not only point towards neural networks as a viable infrastructure for music generation, but also towards training such a system in a manner that is (loosely) more akin to the way a human being would. By this I exclusively refer to the quantity of examples involved in the generative process. More importantly, these results outline the possibility of integrating generative networks and few-shot learning into the production process of a musician. To be more precise, \textcite{larochelle_few-shot_2017} indicate that instead of training a model on various types of music to influence the style of the generated samples, a musician could substitute the data from a support set with music of their own choice and obtain a new generative model for that particular style of music (admittedly, after a few more hours of training).

@@

This section aims to provide a clear and concise answer to the research questions proposed in this Chapter \ref{chap:intro}. They were formulated as follows:

The answers to both questions are summarized in Table \ref{tab:results}. Specifically, the evaluation of the Reptile models indicates that samples from these models are not exactly as strong as those of the baseline. However, the majority of the Reptile samples seem to lag only slightly, reason for which they could be aptly described as competitive and functionally equivalent.

With regards to the performances of expert pianists, the Reptile samples failed to reach the same levels of mastery. However, it should be noted that the baseline showed similar shortcomings. It was hypothesized that the reason for this is the short training which the models had undergone. In this light, a tentative answer to the second research question is that the Reptile models can be similar to real music. Here, a significant area of improvement lies in embedding a sense of motif into the models (since this is what the models have show to lack the most).

\sloppy
\printbibliography

### 2019-05-10 plots/performance_rnn_1_5beethoven-transposed.tex
first draft done and added baseline_50

\begin{axis}[
height=\heatheight,
tick align=outside,
tick pos=left,
width=\heatwidth,
x grid style={white!69.01960784313725!black},
xlabel={Events Sequences},
xmin=-0.5, xmax=199.5,
xtick style={color=black},
y grid style={white!69.01960784313725!black},
ylabel={Activations},
ymin=-0.5, ymax=511.5,
ytick style={color=black}
]
\addplot graphics [includegraphics cmd=\pgfimage,xmin=-0.5, xmax=199.5, ymin=511.5, ymax=-0.5] {images/performance_rnn/performance_rnn_1_5enescu-transposed-000.png};
\end{axis}

>>>>

\begin{axis}[
height=\heatheight,
tick align=outside,
tick pos=left,
width=\heatwidth,
x grid style={white!69.01960784313725!black},
xlabel={Events Sequences},
xmin=-0.5, xmax=199.5,
xtick style={color=black},
y grid style={white!69.01960784313725!black},
ylabel={Activations},
ymin=-0.5, ymax=511.5,
ytick style={color=black}
]
\addplot graphics [includegraphics cmd=\pgfimage,xmin=-0.5, xmax=199.5, ymin=511.5, ymax=-0.5] {performance_rnn_1_5beethoven-transposed-000.png};
\end{axis}
